---
title: "Lab 07 - Conveying the right message through visualisation"
author: "Verity Elliott"
date: "02_16_2026"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

### Exercise 1

The primary limitation of the original figure is its use of dual y-axes with different numerical scales to display mask and non-mask counties. The mask counties are plotted on a scale of approximately 15–26 cases per 100,000, whereas the non-mask counties are plotted on a separate 4–14 scale. Because each series is scaled independently, the visual slopes and relative distances between points are not directly comparable across groups. This violates the principle of common scaling in comparative graphics and increases the risk that viewers will misinterpret similarity in visual pattern as similarity in magnitude or rate of change.

A secondary issue (that irritated me to no end while working on this) concerns the x-axis formatting. The date labels appear offset from their corresponding tick marks, which reduces temporal precision and makes day-specific comparisons more difficult. Although less consequential than the scaling issue, this misalignment further diminishes interpretability.

```{r create_data_frame}
covid_data <- tribble(
  ~date,        ~county_type, ~cases_per_100k,
  "7/12/2020",  "Mask",       25.3,
  "7/12/2020",  "No Mask",    9.8,
  "7/13/2020",  "Mask",       19.8,
  "7/13/2020",  "No Mask",    9.2,
  "7/14/2020",  "Mask",       19.7,
  "7/14/2020",  "No Mask",    9.4,
  "7/15/2020",  "Mask",       20.5,
  "7/15/2020",  "No Mask",    9.8,
  "7/16/2020",  "Mask",       19.8,
  "7/16/2020",  "No Mask",    9.9,
  "7/17/2020",  "Mask",       19.8,
  "7/17/2020",  "No Mask",    9.5,
  "7/18/2020",  "Mask",       20.4,
  "7/18/2020",  "No Mask",    9.5,
  "7/19/2020",  "Mask",       19.9,
  "7/19/2020",  "No Mask",    9.1,
  "7/20/2020",  "Mask",       20.7,
  "7/20/2020",  "No Mask",    8.7,
  "7/21/2020",  "Mask",       21.4,
  "7/21/2020",  "No Mask",    8.7,
  "7/22/2020",  "Mask",       19.9,
  "7/22/2020",  "No Mask",    8.8,
  "7/23/2020",  "Mask",       19.9,
  "7/23/2020",  "No Mask",    8.7,
  "7/24/2020",  "Mask",       20.5,
  "7/24/2020",  "No Mask",    9.8,
  "7/25/2020",  "Mask",       19.0,
  "7/25/2020",  "No Mask",    9.9,
  "7/26/2020",  "Mask",       19.6,
  "7/26/2020",  "No Mask",    10.1,
  "7/27/2020",  "Mask",       17.0,
  "7/27/2020",  "No Mask",    9.6,
  "7/28/2020",  "Mask",       16.2,
  "7/28/2020",  "No Mask",    9.6,
  "7/29/2020",  "Mask",       16.4,
  "7/29/2020",  "No Mask",    9.7,
  "7/30/2020",  "Mask",       16.5,
  "7/30/2020",  "No Mask",    10.0,
  "7/31/2020",  "Mask",       16.0,
  "7/31/2020",  "No Mask",    9.0,
  "8/01/2020",  "Mask",       16.1,
  "8/01/2020",  "No Mask",    9.2,
  "8/02/2020",  "Mask",       15.8,
  "8/02/2020",  "No Mask",    9.0,
  "8/03/2020",  "Mask",       15.9,
  "8/03/2020",  "No Mask",    9.2
)

covid_data <- covid_data %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y"))
```

### Exercise 2

The revised figure improves interpretability by plotting both mask and non-mask counties on a single, common y-axis. By using one shared scale for cases per 100,000, the slopes and vertical distances between the lines are directly comparable across groups. This adheres to the principle of common scaling in comparative data visualization and reduces the risk of perceptual distortion introduced by dual-axis graphics. Differences in magnitude and change over time can therefore be evaluated visually without ambiguity.

In addition, the use of a single x-axis tied directly to the date variable ensures that tick marks and labels correspond clearly to specific time points. Combined with clear axis labels and a unified legend distinguishing county type by color, the overall design improves transparency, comparability, and interpretive accuracy relative to the original figure.

```{r honest_plot}
ggplot(covid_data, aes(x = date, y = cases_per_100k, color = county_type)) +
  geom_line(size = 1) +
  labs(
    title = "Kansas COVID-19 7-Day Average Cases per 100K",
    x = "Date",
    y = "Cases per 100K",
    color = "County Type"
  ) +
  scale_x_date(
    limits = c(as.Date("2020-07-12"), as.Date("2020-08-03")),
    date_breaks = "3 days",
    date_labels = "%m/%d/%Y"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Exercise 3

The revised visualization makes it clearer that mask and non-mask counties experienced substantially different COVID-19 case levels and trends over the study period. Mask counties had a higher average 7-day case rate (mean = 19.0 per 100K) compared with non-mask counties (mean = 9.4 per 100K), but cases in mask counties declined more sharply over time (−9.4 per 100K) than in non-mask counties (−0.6 per 100K). This pattern of relative magnitude and temporal change is evident in the plot, whereas the original dual-axis graph obscured these differences by using independent y-axes that distorted the visual comparison.

```{r summary_statistics}
library(dplyr)

# Summary statistics by county type
covid_summary <- covid_data %>%
  group_by(county_type) %>%
  summarise(
    mean_cases = mean(cases_per_100k),
    min_cases = min(cases_per_100k),
    max_cases = max(cases_per_100k),
    sd_cases = sd(cases_per_100k),
    first_day = first(cases_per_100k),
    last_day = last(cases_per_100k),
    change = last_day - first_day
  )

print(covid_summary)
```

### Exercise 4
